---
title: 'UMCCR Transcriptome Patient Summary'
author: 'Jacek Marzec, Sehrish Kanwal, Lavinia Gordon'
date: '`r Sys.Date()`'
output:
  html_document:
    theme: readable
    toc: true
    toc_float: true
    code_folding: hide
  rmdformats::material:
    highlight: kate
params:
  report_dir: '../reports'
  sample_name: 'CCR170012_MH17T001P013'
  count_file: '../data/CCR170012_MH17T001P013-ready.counts'
  datasets_pancreas: '../data/Datasets_list_pancreas.txt'
  datasets_bailey: '../data/Datasets_list_bailey.txt'
  datasets_collisson: '../data/Datasets_list_collisson.txt'
  datasets_moffitt: '../data/Datasets_list_moffitt.txt'
  genes_selected: '../data/Selected_genes_ensmebl.txt'
---

```{r define_functions, comment=NA, message=FALSE, warning=FALSE}
##### Define functions
##### Create 'not in' operator
"%!in%" <- function(x,table) match(x,table, nomatch = 0) == 0

##### Prepare object to write into a file
prepare2write <- function (x) {
  
  x2write <- cbind(rownames(x), x)
  colnames(x2write) <- c("Gene",colnames(x))
  return(x2write)
}

##### Combine sample expression profile with reference datasets. This function outputs a vector with first element containing the merged data and second element containing merged targets info
combineDatasets <- function(sample_name, sample_counts, ref_dataset) {
  
  ##### Read file with reference datasets information
  DatasetInput=read.table(ref_dataset, sep="\t", as.is=TRUE, header=TRUE, row.names=1)
  
  ##### Extract info about target file for the first dataset
  fileInfo = strsplit(DatasetInput[,"Target_file"], split='/', fixed=TRUE)
  targetFile <- read.table(DatasetInput[1,"Target_file"], sep="\t", as.is=TRUE, header=TRUE)[,c(1:4)]
  rownames(targetFile) <- targetFile[,"Sample_name"]
  targetFile <- cbind(targetFile[,2:4],rownames(DatasetInput[1,]))
  colnames(targetFile)[ncol(targetFile)] <- "Dataset"
  
  if ( nrow(DatasetInput) > 1 ) {
    for ( i in 2:nrow(DatasetInput) ) {
        
      ##### Create a temporary object to store info from the remaining target files
      targetFileTmp <- read.table(DatasetInput[i,"Target_file"], sep="\t", as.is=TRUE, header=TRUE)[,c(1:4)]
      rownames(targetFileTmp) <- targetFileTmp[,"Sample_name"]
      targetFileTmp <- cbind(targetFileTmp[,2:4],rownames(DatasetInput[i,]))
      colnames(targetFileTmp)[ncol(targetFileTmp)] <- "Dataset"
        
      targetFile <- rbind(targetFile, targetFileTmp)
    }
  }  
  
  ##### Add sample info
  sampleTargetFile <- data.frame(sample_counts, sample_name, NA, sample_name)
  names(sampleTargetFile) <- names(targetFile)
  rownames(sampleTargetFile) <- sample_name
  targetFile <- rbind( targetFile, sampleTargetFile )
  
  ##### Make syntactically valid names
  rownames(targetFile) <- make.names(rownames(targetFile))
  
  ##### Read sample read count file and combine it with reference datasets
  datasets.comb=read.table(sample_counts, sep="\t", as.is=TRUE, header=FALSE, row.names=NULL)
  names(datasets.comb) <- c("", sample_name)
      
  ##### list genes present in the read count file
  gene_list <- as.vector(datasets.comb[,1])
      
  ##### Loop through the expression data from different datasets and merge them into one matrix
  for ( data_matrix in DatasetInput[ , "Expression_matrix" ] ) {
    
    ##### Add data from the reference datasets
    dataset <- as.data.frame( read.table(data_matrix, header=TRUE, sep="\t", row.names=NULL) )
      
    ##### list genes present in individal files
    gene_list <- c( gene_list, as.vector(dataset[,1]) )
    
    ##### Merge the expression datasets and make sure that the genes order is the same
    datasets.comb <- merge( datasets.comb, dataset, by=1, all = FALSE, sort= TRUE)
      
    ##### Remove per-sample data for merged samples to free some memory
    rm(dataset)
  }
  
  ##### Use gene IDs as rownames
  rownames(datasets.comb) <- datasets.comb[,1]
  datasets.comb <- datasets.comb[, -1]
  
  ##### Make syntactically valid names
  colnames(datasets.comb) <- make.names(colnames(datasets.comb))
  
  ##### Make sure that the target file contains info only about samples present in the data matrix
  targetFile <- targetFile[ rownames(targetFile) %in% colnames(datasets.comb),  ]
  
  ##### Make sure that the samples order in the data matrix is the same as in the target file 
  datasets.comb <- datasets.comb[ , rownames(targetFile) ]
  
  ##### Identify genes that were not present across all per-sampel files and were ommited in the merged matrix
  gene_list <- unique(gene_list)
  gene_list.missing <- gene_list[ gene_list %!in% rownames(datasets.comb) ]
  
  ##### Write list of missing genes into a file
  if ( length(gene_list.missing) > 0 ) {
    write.table(prepare2write(gene_list.missing), file = paste0(params$report_dir, "/", sample_name,".missing_genes.txt"), sep="\t", quote=FALSE, row.names=TRUE, append = FALSE )
  }
  
    return( list(datasets.comb, targetFile) )
}


##### Assign colours to different groups
getTargetsColours <- function(targets) {
  
##### Predefined selection of colours for groups
targets.colours <- c("red","blue","green","darkgoldenrod","darkred","deepskyblue", "coral", "cornflowerblue", "chartreuse4", "bisque4", "chocolate3", "cadetblue3", "darkslategrey", "lightgoldenrod4", "mediumpurple4", "orangered3","indianred1","blueviolet","darkolivegreen4","darkgoldenrod4","firebrick3","deepskyblue4", "coral3", "dodgerblue1", "chartreuse3", "bisque3", "chocolate4", "cadetblue", "darkslategray4", "lightgoldenrod3", "mediumpurple3", "orangered1")
  
  f.targets <- factor(targets)
  vec.targets <- targets.colours[1:length(levels(f.targets))]
  targets.colour <- rep(0,length(f.targets))
  for(i in 1:length(f.targets))
    targets.colour[i] <- vec.targets[ f.targets[i]==levels(f.targets)]
  
  return( list(vec.targets, targets.colour) )
}

##### Assign colours to different datasets
getDatasetsColours <- function(datasets) {
  
  ##### Predefined selection of colours for datasets
  datasets.colours <- c("dodgerblue","firebrick","lightslategrey","darkseagreen","orange","darkcyan","bisque", "coral2", "cadetblue3","red","blue","green")
  
  f.datasets <- factor(datasets)
  vec.datasets <- datasets.colours[1:length(levels(f.datasets))]
  datasets.colour <- rep(0,length(f.datasets))
  for(i in 1:length(f.datasets))
    datasets.colour[i] <- vec.datasets[ f.datasets[i]==levels(f.datasets)]
  
  return( list(vec.datasets, datasets.colour) )
}

##### Perform PCA. This function outputs a list with dataframe and samples colouring info ready for plotting
pca <- function(data, targets) {

  ##### Keep only genes with variance > 0 across all samples
  rsd <- apply(data,1,sd)
  data.subset <- data[rsd>0,]
  
  ##### Perform PCA
  data.subset_pca <- prcomp(t(data.subset), scale=FALSE)
  
  ##### Get variance importance for all principal components
  importance_pca <- summary(data.subset_pca)$importance[2,]
  importance_pca <- paste(round(100*importance_pca, 2), "%", sep="")
  names(importance_pca) <- names(summary(data.subset_pca)$importance[2,])
    
  ##### Prepare data frame
  data.subset_pca.df <- data.frame(targets$Target, targets$Dataset, data.subset_pca$x[,"PC1"], data.subset_pca$x[,"PC2"], data.subset_pca$x[,"PC3"])
  colnames(data.subset_pca.df) <- c("Target", "Dataset", "PC1", "PC2", "PC3")
  
  ##### Assigne colours to targets and datasets
  targets.colour <- getTargetsColours(target$Target)
  datasets.colour <- getDatasetsColours(target$Dataset)
  
  ##### Create a list with dataframe and samples colouring info
  pca.list <- list(data.subset_pca.df, importance_pca, targets.colour, datasets.colour)
  names(pca.list) <- c("pca.df", "importance_pca", "targets", "datasets")
  
  return( pca.list )
}
  

##### Generate cumulative distribution function (CDF) plot for selected gene using sample combined with pancreas expression data
cdfPlot <- function(gene, data, targets, sampleName) {
  
  ##### Subset data for each biological group
  sample.expr <- sort(data[, sampleName])
  normal.expr <- sort(data[, "SRR1464226"])
  PDAC.expr <- sort(data[, "TCGA.IB.AAUO.01A.12R.A38C.07"])
  met.expr <- sort(data[, "TCGA.HZ.A9TJ.06A.11R.A41B.07"])
  
  ##### Perform range standardization between 0 and 1 (for the cumulative sums)
  sample.expr <- sort(sample.expr-min(sample.expr))/(max(sample.expr)-min(sample.expr))
  normal.expr <- sort(normal.expr-min(normal.expr))/(max(normal.expr)-min(normal.expr))
  PDAC.expr <- sort(PDAC.expr-min(PDAC.expr))/(max(PDAC.expr)-min(PDAC.expr))
  met.expr <- sort(met.expr-min(met.expr))/(max(met.expr)-min(met.expr))
  
  ##### Calculate cumulative sums and organise the data into data frame
  sample.expr.cum <- as.data.frame(cbind(c(1:length(sample.expr)), sample.expr, cumsum(sample.expr)))
  normal.expr.cum <- as.data.frame(cbind(c(1:length(normal.expr)), normal.expr, cumsum(normal.expr)))
  PDAC.expr.cum <- as.data.frame(cbind(c(1:length(PDAC.expr)), PDAC.expr, cumsum(PDAC.expr)))
  met.expr.cum <- as.data.frame(cbind(c(1:length(met.expr)), met.expr, cumsum(met.expr)))
  
  names(sample.expr.cum) <- c("index", "expression", "cumulative_fraction")
  names(normal.expr.cum) <- c("index", "expression", "cumulative_fraction")
  names(PDAC.expr.cum) <- c("index", "expression", "cumulative_fraction")
  names(met.expr.cum) <- c("index", "expression", "cumulative_fraction")
  
  ##### Extract expression for selected genes
  sample.expr.cum.selected <- sample.expr.cum[ rownames(sample.expr.cum) %in% unlist(gene), ]
  normal.expr.cum.selected <- normal.expr.cum[ rownames(normal.expr.cum) %in% unlist(gene), ]
  PDAC.expr.cum.selected <- PDAC.expr.cum[ rownames(PDAC.expr.cum) %in% unlist(gene), ]
  met.expr.cum.selected <- met.expr.cum[ rownames(met.expr.cum) %in% unlist(gene), ]
  
  ##### Generate interactive CFD plot with plotly
  p <- plot_ly(sample.expr.cum, x = ~expression, color = I("black")) %>%
  
    ##### Add sample data
    add_markers(y = sample.expr.cum.selected$index, x = sample.expr.cum.selected$expression,
                text = rownames(sample.expr.cum.selected),
                name = "Selected genes",
                marker = list(size = 9, color = "black"),
                showlegend = FALSE) %>%
  
    add_lines(y = sample.expr.cum$index, x = sample.expr.cum$expression, 
              line = list(color = "grey"),
              text = rownames( met.expr.cum.selected),
              name = params$sample, showlegend = TRUE) %>%
    
    ##### Add normal pancreas data
    add_markers(y = normal.expr.cum.selected$index, x =  normal.expr.cum.selected$expression,
                text = rownames( normal.expr.cum.selected),
                name = "Selected genes",
                marker = list(size = 9, color = "green"),
                showlegend = FALSE) %>%
  
    add_lines(y = normal.expr.cum$index, x = normal.expr.cum$expression, opacity = 0.5,
              line = list(color = "green", dash = "dash"),
              text = rownames( met.expr.cum.selected),
              name = "Pancreas (normal)", showlegend = TRUE) %>%
    
    ##### Add PDAC data
    add_markers(y = PDAC.expr.cum.selected$index, x =  PDAC.expr.cum.selected$expression,
                text = rownames( PDAC.expr.cum.selected),
                name = "Selected genes",
                marker = list(size = 9, color = "coral"),
                showlegend = FALSE) %>%
  
    add_lines(y = PDAC.expr.cum$index, x = PDAC.expr.cum$expression, opacity = 0.5,
              line = list(color = "coral", dash = "dash"),
              text = rownames( met.expr.cum.selected),
              name = "PDAC", showlegend = TRUE) %>%
    
    ##### Add metastatic PDAC data
    add_markers(y = met.expr.cum.selected$index, x =  met.expr.cum.selected$expression,
                text = rownames( met.expr.cum.selected),
                name = "Selected genes",
                marker = list(size = 9, color = "red"),
                showlegend = FALSE) %>%
    
    add_lines(y = met.expr.cum$index, x = met.expr.cum$expression, opacity = 0.5,
              line = list(color = "red", dash = "dash"),
              text = rownames( met.expr.cum.selected),
              name = "PDAC (metastatic)", showlegend = TRUE) %>%
  
    layout(xaxis = list(title = "mRNA expression"),
           yaxis = list(title = "Cumulative fraction"),
           legend = list(orientation = 'v', x = 0.02, y = 0.9, bgcolor = "transparent"),
           annotations = list(x = sample.expr.cum.selected$expression, y = sample.expr.cum.selected$index,
    text = rownames(sample.expr.cum.selected), xref = "x", yref = "y", showarrow = TRUE, arrowhead = 0, opacity = 0.5, ax = 100, ay = 100
  ))
  return( p )
}

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot <- function(gene, data, targets, targets.colour, sampleName, order ) {
  
  ##### Extract expression for selected genes
  gene.expr <- data[ rownames(data) %in% unlist(gene), ]

  ##### Prepare data frame
  gene.expr.df <- data.frame(targets, as.numeric(gene.expr))
  colnames(gene.expr.df) <- c("Group", "Expression")
  
  ##### Apply defined group order
  order <- order[order %in% gene.expr.df$Group]
  gene.expr.df$Group <- factor(gene.expr.df$Group, levels=order)
  
  p <- plot_ly(gene.expr.df, y= ~Expression, color = ~Group, type = 'box', jitter = 0.3, pointpos = 0, boxpoints = 'all', colors = targets.colour[[1]], width = 800, height = 600) %>%
  layout(yaxis = list( title ="mRNA expression"), margin = list(l=50, r=50, b=50, t=50, pad=4), autosize = F, legend = list(orientation = 'v', y = 0.5), showlegend=TRUE)
  
  return( p )
}
```

```{r load_libraries, warning=FALSE}
##### Load libraries
suppressMessages(library(edgeR))
suppressMessages(library(preprocessCore))
suppressMessages(library(rapportools))
suppressMessages(library(plotly))
suppressMessages(library(edgeR))
```


```{r load_ref_data, message=FALSE, warning=FALSE}
##### Load reference datasets
##### Create a list with reference datasets
ref_datasets <- c("pancreas", "bailey", "collisson", "moffitt")
ref_datasets.list <- vector("list", length(ref_datasets))
names(ref_datasets.list) <- ref_datasets

##### Create a list with subtype-associated genes
ref_genes <- c("selected")
ref_genes.list <- vector("list", length(ref_genes))
names(ref_genes.list) <- ref_genes

##### Read in reference datasets and merge them with sample data. This part outputs a vector with first element containing the merged data and second element containing merged targets info
ref_datasets.list[["pancreas"]] <- combineDatasets(params$sample_name, params$count_file, params$datasets_pancreas)
ref_datasets.list[["bailey"]] <- combineDatasets(params$sample_name, params$count_file, params$datasets_bailey)
ref_datasets.list[["collisson"]] <- combineDatasets(params$sample_name, params$count_file, params$datasets_collisson)
ref_datasets.list[["moffitt"]] <- combineDatasets(params$sample_name, params$count_file, params$datasets_moffitt)

##### Read in selected genes list
ref_genes.list[["selected"]] <- read.table(params$genes_selected, sep="\t", as.is=TRUE, header=FALSE, row.names=NULL)
```

```{r data_transformation, comment = NA, message=FALSE, warning=FALSE}
##### Data transformation and filtering
##### For differential expression and related analyses, gene expression is rarely considered at the level of raw counts since libraries sequenced at a greater depth will result in higher counts. Rather, it is common practice to transform raw counts onto a scale that accounts for such library size differences. Here we convert the read count data into log2-counts per million (***log-CPM***) using function from *[edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html)* package. Genes with very low counts across all libraries provide little evidence for differential expression. In the biological point of view, a gene must be expressed at some minimal level before it is likely to be translated into a protein or to be biologically important. In addition, the pronounced discretenes of these counts interferes with some of the statistical approximations that are used later in the pipeline. These genes should be filtered out prior to further analysis.

##### Loop through combined datasets
for ( ref in names(ref_datasets.list) ) {
  
  counts <- ref_datasets.list[[ref]][[1]]
  target <- ref_datasets.list[[ref]][[2]]
  
  ##### Create EdgeR DGEList object
  y <- DGEList(counts=counts,  group=target$Target)
  
  ##### Add datasets name for each sample
  y$samples$dataset <- target$Dataset
  
  ##### Filtering to remove low expressed genes. Users should filter with CPM rather than filtering on the counts directly, as the latter does not account for differences in library sizes between samples. Here we keep only genes that have CPM of 1
  keep <- rowSums(cpm(y)>1) >= ncol(counts)/10
  y.filtered <- y[keep, , keep.lib.sizes=FALSE]
  
  ref_datasets.list[[ref]][[3]] <- y.filtered
}

# cat("The CPM of 1 (cut-off for removing low expressed genes) corresponds to", round(min(as.numeric(colSums(counts)*1e-6)), digits=0), "reads in sample with the lowest sequencing depth, and", round(max(as.numeric(colSums(counts)*1e-6)), digits=0), "reads in sample with the greatest sequencing depth\n")

cat(nrow(y.filtered$counts), "genes remained after filtering low expressed genes, out of the total", nrow(counts), "input genes\n\n")
```


```{r data_normalisation, comment = NA, message=FALSE, warning=FALSE}
##### Data normalisation
##### During the sample preparation or sequencing process, external factors that are not of biological interest can affect the expression of individual samples. For example, samples processed in the first batch of an experiment can have higher expression overall when compared to samples processed in a second batch. It is assumed that all samples should have a similar range and distribution of expression values. Normalisation for sample-specific effectss is required to ensure that the expression distributions of each sample are similar across the entire experiment. Normalisation by the method of *[trimmed mean of M-values](https://www.ncbi.nlm.nih.gov/pubmed/20196867) (TMM)* is performed using the *calcNormFactors* function in *[edgeR](https://bioconductor.org/packages/release/bioc/html/edgeR.html)*. The normalisation factors calculated here are used as a scaling factor for the library sizes. TMM is the recommended for most RNA-Seq data where the majority (more than half) of the genes are believed not differentially expressed between any pair of the samples.

##### Adjust for RNA composition effect. Calculate scaling factors for the library sizes with calcNormFactors function using trimmed mean of M-values (TMM) between each pair of samples. Note, that the raw read counts are used to calculate the normalisation factors

##### Loop through combined datasets
for ( ref in names(ref_datasets.list) ) {
  
  y.filtered <- ref_datasets.list[[ref]][[3]]
  
  y.filtered.norm <- calcNormFactors(y.filtered, method = "TMM")
  
  ##### Transformations from the raw-scale to CPM
  y.filtered.norm.cpm <- cpm(y.filtered.norm, normalized.lib.sizes=TRUE, log=TRUE, prior.count=0.25)
  
  ref_datasets.list[[ref]][[3]] <- y.filtered.norm.cpm
}
```

```{r pca, comment = NA, message=FALSE, warning=FALSE}
##### Principal component analysis (PCA)
##### Loop through combined datasets and perform PCA
for ( ref in names(ref_datasets.list) ) {
  
  target <- ref_datasets.list[[ref]][[2]]
  data <- ref_datasets.list[[ref]][[3]]
  
  ref_datasets.list[[ref]][[4]] <- pca(data, target)
}
```

## Seleced genes expression  {.tabset}

This section presents the...

### `r ref_genes.list[["selected"]][1,]`

```{r cdf_plot_pancreas_1, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][1,]

cdfPlot(gene, data, target, params$sample_name)

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][2,]`

```{r cdf_plot_pancreas_2, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][2,]

cdfPlot(gene, data, target, params$sample_name)

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][3,]`

```{r cdf_plot_pancreas_3, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][3,]

cdfPlot(gene, data, target, params$sample_name)

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][4,]`

```{r cdf_plot_pancreas_4, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][4,]

cdfPlot(gene, data, target, params$sample_name)

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][5,]`

```{r cdf_plot_pancreas_5, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][5,]

cdfPlot(gene, data, target, params$sample_name)

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

## Seleced genes expression profile {.tabset}

The expression profile of **XXX** genes across pancreas-related lesions/tissues are presented. The box-plot demonstrates mRNA expression level quartiles (y-axis) in samples stratified according to their group (x-axis). 

### `r ref_genes.list[["selected"]][1,]`

```{r box_plot_pancreas_1, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
target <- target$Target
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][1,]

order = c(params$sample_name, "Pancreas (normal)", "Pancreas (normal adjacent)", "IPMN", "PanNET", "PDAC", "PDAC (metastatic)")

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot(gene, data, target, targets.colour, params$sample_name, order )
  
##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][2,]`

```{r box_plot_pancreas_2, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
target <- target$Target
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][2,]

order = c(params$sample_name, "Pancreas (normal)", "Pancreas (normal adjacent)", "IPMN", "PanNET", "PDAC", "PDAC (metastatic)")

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot(gene, data, target, targets.colour, params$sample_name, order )

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][3,]`

```{r box_plot_pancreas_3, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
target <- target$Target
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][3,]

order = c(params$sample_name, "Pancreas (normal)", "Pancreas (normal adjacent)", "IPMN", "PanNET", "PDAC", "PDAC (metastatic)")

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot(gene, data, target, targets.colour, params$sample_name, order )

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][4,]`

```{r box_plot_pancreas_4, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
target <- target$Target
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets
gene <- ref_genes.list[["selected"]][4,]

order = c(params$sample_name, "Pancreas (normal)", "Pancreas (normal adjacent)", "IPMN", "PanNET", "PDAC", "PDAC (metastatic)")

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot(gene, data, target, targets.colour, params$sample_name, order )

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

### `r ref_genes.list[["selected"]][5,]`

```{r box_plot_pancreas_5, comment = NA, message=FALSE, warning=FALSE, fig.width = 7, fig.height = 4}
suppressMessages(library(plotly))
##### Generate PCA plots for pancreatic related lesions
ref <- "pancreas"

target <- ref_datasets.list[[ref]][[2]]
target <- target$Target
data <- ref_datasets.list[[ref]][[3]]
targets.colour <- ref_datasets.list[[ref]][[4]]$targets
datasets.colour <- ref_datasets.list[[ref]][[4]]$datasets

gene <- ref_genes.list[["selected"]][5,]

order = c(params$sample_name, "Pancreas (normal)", "Pancreas (normal adjacent)", "IPMN", "PanNET", "PDAC", "PDAC (metastatic)")

##### Generate box-plot for selected gene using sample combined with pancreas expression data
boxPlot(gene, data, target, targets.colour, params$sample_name, order )

##### Detach plotly package. Otherwise it clashes with other graphics devices
detach("package:plotly", unload=FALSE)
```

## Parameters

```{r params_info, comment = NA}
for ( i in 1:length(params) ) {

  cat(paste("Parameter: ", names(params)[i], "\nValue: ", paste(unlist(params[i]), collapse = ","), "\n\n", sep=""))
}
```

## Addendum

```{r sessioninfo}
devtools::session_info()
```

