---
title: "Fusion Results"
author: "Sehrish Kanwal"
date: "01/10/2018"
output: html_document
params:
  pizzly_folder: '/Users/kanwals/Documents/UMCCR/data/fusions/pizzly-validation'
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

### Required packages

```{r}
suppressMessages(library(kableExtra))
suppressMessages(library(tidyverse))
suppressMessages(library(knitr))
suppressMessages(library(magick))
```

### Prioritization of Pizzly results

The prioritization of pizzly results is performed in two steps.

1. In the first step, basic idea is to run kallisto to quantify the fusion transcripts (reported by pizzly) and 
select those which have a decent TPM support.
  * Create a new index based on the transcriptome and the fusion transcripts identified by pizzly.
  * Run kallisto in normal quantification mode on the expanded index to quantify both normal transcripts 
  and fusions.
  * Select only the fusions with a high TPM value as reported by previous step.

2. In the second step, fusion genes reported by pizzly are compared with PMCC gene list and report 
the ones that are present in this list. 

```{r}
#read in the pizzly fusion calls
pizzly.fusions <- read.table(file = paste(params$pizzly_folder, "pizzly-flat-filtered.tsv", sep = "/"), header = TRUE)

#read in the transcripts quantification file
quant <- read.table(file = paste(params$pizzly_folder, "abundance.tsv", sep = "/"), header = TRUE)

#sort and filter quantification file on tpm values. Currently filtering on quantiles. Selected 0.997 because that reduces the final fusion
#calls to the value we are interested in (~15)
quant.sorted.filtered <- filter(arrange(quant, desc(quant$tpm)), tpm >= (quantile(quant$tpm, 0.999)))

#initialize an empty dataframe
result <- data.frame()
result.list <- vector("list", length = nrow(pizzly.fusions))

#let's try using for loop for iterating over pizzly.fusions dataframe and get transcriptID and fusion gene pair information.
#can also filter quant.sorted.filtered$target_id to have only fusion gene target ids (that is two tracscripts instead of one-
#this will increase speed

for (row in 1:nrow(pizzly.fusions)){
  y <- strsplit(as.character(pizzly.fusions[row, 7]), "\\;")
  y <- unname(y)
  for (i in 1:length(y[[1]])){
    if (y[[1]][i] %in% quant.sorted.filtered$target_id){
      #val <- c(y[[1]][i], as.character(pizzly.fusions[row,1]), as.character(pizzly.fusions[row, 3]))
      #creating a new dataframe for the filtered pizzly results
      result <- rbind(result, data.frame(pizzly.fusions[row,]))
      #result.list[[row]] <- val
    }
  }
}

#remove duplicated values from result filtered using expression count(as multiple transcripts might support fusion between same gene)
deduped.result <- unique(result)

#Extract only those fusion genes that are in PMCC list
#read in the PMCC genes list
pmcc.genes <- read.table(file = '~/Documents/UMCCR/data/RNAseq-report/2018-02-16_PMCC_Panel_Genes.txt', header = FALSE)
result.pmcc <- data.frame()
for (row in 1:nrow(pizzly.fusions)){
  if(pizzly.fusions[row,1] %in% pmcc.genes$V1 | pizzly.fusions[row,3] %in% pmcc.genes$V1){
    #creating a new dataframe for extracting pizzly rows with pmcc gene hits
    result.pmcc <- rbind(result.pmcc, data.frame(pizzly.fusions[row,]))
  }
}

#sorting pizzly.fusions according to two new target dataframes identified i.e. 1- pizzly result filtering using expressions counts and 2- pmcc gene list

pizzly.fusions$order <- rownames(pizzly.fusions) %in% rownames(deduped.result) 
pizzly.fusions$order2 <- rownames(pizzly.fusions) %in% rownames(result.pmcc)

pizzly.fusions2 <- arrange(pizzly.fusions, -order, -order2) %>%
  select(-order,-order2)

kable(pizzly.fusions2) %>%
  kable_styling(font_size = 10, "striped", "bordered") %>%
  #row_spec(c(as.numeric(rownames(deduped.result)),as.numeric(rownames(result.pmcc))), bold = T, color = "white", background = "#D7261E") %>%
  #row_spec(1:nrow(deduped.result), bold = T, color = "white", background = "red") %>%
  #row_spec((nrow(deduped.result)+1):(nrow(deduped.result) + nrow(result.pmcc)), color = "white", background = "blue") %>%
  group_rows("Filtered results using expression count matrix", 1, nrow(deduped.result), label_row_css = "background-color: #E74C3C; color: #fff;") %>%
  group_rows("Filtered results using PMCC gene list", (nrow(deduped.result)+1), (nrow(deduped.result) + nrow(result.pmcc)), label_row_css = "background-color: #2980B9; color: #fff;") %>%
  group_rows("Other pizzly results", (nrow(deduped.result) + nrow(result.pmcc) +1), nrow(pizzly.fusions2), label_row_css = "background-color: #27AE60; color: #fff;") %>%
  scroll_box(width = "100%", height = "200px")
```

## Fusion visualization {.tabset}

Currently, dispalying the top 3 hits from pizzly filtering output (Group 1).

### `r paste(pizzly.fusions2$geneA.name[1], pizzly.fusions2$geneB.name[1], sep="-")`

```{r fusions_figs1, comment = NA, message=TRUE, warning=FALSE}
#get path to fusion visualization pdf file
fusion_1 <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[1], "_", pizzly.fusions2$geneB.name[1], ".pdf"), sep = "/")

#export pdf to png
fusion1 <- image_read_pdf(fusion_1, pages = NULL, density = 300)
fusion1 <- image_write(fusion1, path = "~/Documents/UMCCR/data/fusions/pizzly-validation/ALB_APOA1.png", format = "png")

#present the converted file in the report
fusion1_png <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[1], "_", pizzly.fusions2$geneB.name[1], ".png"), sep = "/")
include_graphics(fusion1_png)
```

### `r paste(pizzly.fusions2$geneA.name[2], pizzly.fusions2$geneB.name[2], sep="-")`

```{r fusions_figs2, comment = NA, message=TRUE, warning=FALSE}

fusion_2 <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[2], "_", pizzly.fusions2$geneB.name[2], ".pdf"), sep = "/")

#export pdf to png
fusion2 <- image_read_pdf(fusion_2, pages = NULL, density = 300)
image_write(fusion2, path ="~/Documents/UMCCR/data/fusions/pizzly-validation/ALB_FGG.png", format = "png")

#present the converted file in the report
fusion2_png <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[2], "_", pizzly.fusions2$geneB.name[2], ".png"), sep = "/")
include_graphics(fusion2_png)
```

### `r paste(pizzly.fusions2$geneA.name[3], pizzly.fusions2$geneB.name[3], sep="-")`
```{r fusions_figs3, comment = NA, message=TRUE, warning=FALSE}
fusion_3 <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[3], "_", pizzly.fusions2$geneB.name[3], ".pdf"), sep = "/")

#export pdf to png
fusion3 <- image_read_pdf(fusion_3, pages = NULL, density = 300)
fusion3 <- image_write(fusion3, path = "~/Documents/UMCCR/data/fusions/pizzly-validation/HP_ALB.png", format = "png")

#present the converted file in the report
fusion3_png <- paste(params$pizzly_folder, paste0(pizzly.fusions2$geneA.name[3], "_", pizzly.fusions2$geneB.name[3], ".png"), sep = "/")
include_graphics(fusion3_png)
```








